Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-21 aistats.tex
Incorporate Frederik's comments

\subsection{Multivariate MLE}
In higher dimensions, $X \sim \cN(0, \mu_*)$, $X\in\real^d$, $T(X) = XX^\top$, and the mean parameter $\mu_*$ is a symmetric positive definite covariance matrix of order $d$, living in dimension $p = \frac{d(d+1)}{2}$. 
Note that we switched the role of $d$ from $\dim(T(\cX))=d$ to $\dim(\cX) =d$ and $\dim(T(\cX))=p$, where $\dim(\cZ)$ is the dimension of the smallest affine subspace containing $\cZ$.

\fdk{I'm worried talking about dimension might be confused with curved EF (less parameters than sufficient statistics). What do you think of
%
\\~\\
%
In higher dimensions, $X \sim \cN(0, \mu_*)$, $X\in\real^d$, $T(X) = XX^\top$, and the mean parameter $\mu_*$ 
is a $d \times d$ symmetric, positive definite covariance matrix with $p = \frac{d(d+1)}{2}$ degrees of freedom.
Note that here $d$ denotes the dimensionality of the data $X$, rather than the dimensionality of the parameters $\mu_*$.
}

@@

 	&\leq p \paren{ \inv{n-d-1} - \inv{2n}}  + \frac{d + 1}{4n(n - d - 1)}\\

@@

 	&< p \frac{n + d +1}{2 n (n-d-1)}  + \frac{d(d + 1)/2}{2dn(n - d - 1)}\\

@@

where $p = d(d+1)/2$ and the last inequality used $\inv{2d}\leq 1$. 

@@

\begin{proof}
	The lower bound can be readily obtained by applying Jensen's inequality 
	to the convex function $x\mapsto\inv{x}$ for $x>0$.
	The upper bound requires more work.
	To start, let us plug in the definition of $\MAPm$
\alignn{
		\expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\nat^*}} 
		= \expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i^2}}  
		= (n_0 + n) \expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} 
		= (n_0 +n) \expect{\inv{a + \chi^2(n)}} \; ,
}
	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and $a=n_0 \frac{\mu_0}{\mu^*}$. The expectation reads
\alignn{
 	\expect{\inv{a + \chi^2(n)}}  
	= \int_0^\infty \inv{a+x} \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{2^{\half[n]}\Gamma(\half[n])} dx \; .
}
	This $1/(a+x)$ is hard to deal with.
	Fortunately, there is a trick to express it as an integral itself , with exponential terms that fit well with the $\chi^2$-density,
	\fdk{I don't follow from there.} \rlp{do you have more detailed feedback ? I don't know how to improve it.}
\alignn{
		&\int_1^\infty e^{-\frac{zt}{2}} dt = \frac{2}{z} e^{-\half[z]} \\
		\implies &\inv{z} = \half e^{\half[z]} \int_1^\infty e^{-\frac{zt}{2}} dt \\
		\implies & \boxed{\inv{a+x} = \half e^{\frac{a+x}{2}} \int_1^\infty e^{- \frac{(a+x)t}{2}}dt  \; .}
} 
We can then plug this integral into the expectation and apply Fubini's theorem to get a tractable integral on the inside
\alignn{
		2^{\half[n]}\Gamma(\half[n]) \expect{\inv{a + \chi^2(n)}}  
		&= \int_0^\infty \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{a+x} dx \\
		&= \int_0^\infty dx\  x^{\frac{n}{2}-1}e^{-\half[x]} \half e^{\half[a+x]} \int_1^\infty  dt\  e^{-\half[(a+x)t]}  \\
		&= \half e^{\half[a]} \int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty dx \  x^{\frac{n}{2}-1} e^{-\half[x t]}  \; ,
}
	assuming we can exchange the order of integration.
	%\fdk{Fubini's theorem would be a justification here, right? Are you trying to say ``assuming we can exchange the order of integration''?}\rlp{yes exactly.}
	Now with the change of variable $x = 2\frac{y}{t}$, eg $dx = 2\frac{dy}{t}$, the inner integral becomes
	\begin{align}
		\expect{\inv{a + \chi^2(n)}}  
		&= \frac{e^{\half[a]} }{2^{\half[n]+1}\Gamma(\half[n]) }\int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty 2\frac{dy}{t} \  (2\frac{y}{t})^{\frac{n}{2}-1} e^{-y} \\
		&= \frac{e^{\half[a]} }{2\Gamma(\half[n]) } 
		\underbrace{\int_1^\infty  \frac{e^{-\half[a t]}}{t^{\frac{n}{2}}}  dt}_{E_{\half[n]}(\half[a])  } \ 
		\underbrace{\int_0^\infty  y^{\frac{n}{2}-1} e^{-y} dy}_{\Gamma(\half[n])}
	\end{align}
	so we finally get  the formula valid for all $n\geq 1$
	\begin{align}
		\boxed{\expect{\inv{a + \chi^2(n)}}  = \half e^{\frac{a}{2}} E_{\half[n]}(\half[a])}
		\label{eq:hard_expectation}
	\end{align}
	where the {generalized exponential integral function} is defined as 
	\begin{align}
		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; .
	\end{align}
	Now our goal is to bound this function with simpler functions.
	Fortunately, mathematicians have been working on these integrals for decades.
	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19.E21}{the general bound} \citep[Eq.~8.19.21]{DLMF}
	\begin{align}
		e^x E_k(x) 
		&\leq \inv{x + k - 1} \\
		\iff \expect{\inv{a + \chi^2(n)}}
		&\leq \inv{a + n - 2} \\
		\iff \expect{\frac{\mu^*}{\MAPm}} 
		&\leq \frac{n +n_0}{a + n - 2}
	\end{align}
We are left with a special case when  $n=1$, eg $k=\half$. 
Then we can use the trivial bound 
\begin{align}
	\expect{\inv{a + X^2}} < \inv{a}
	\label{eq:trivial_bound}
\end{align}
to conclude the proof.
\end{proof}

@@

\begin{definition}[Self-concordance]
\label{def:self-concordance}
A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies 
\alignn{\abs{g'''(t)} \leq 2 g''(t)^{\frac{3}{2}}}
for all feasible $t$.
\end{definition}

\begin{definition}[Pseudo Self-concordance] \citep{bach2010self} 
A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies 
\alignn{\abs{g'''(x)} \leq C \norm{v}_2 g''(x)}
for all feasible $t$ and for some $C\geq 0$.
\end{definition}

@@

We quickly reviews some important properties of self-concordant functions, introduced in \citep{nesterov2003introductory}. We start with a few notations. Let $A^*$ be a self-concordant function. Then, we write
\begin{itemize}
	\item the local norm $\|\cdot\|_x = \sqrt{ \langle \nabla^2A^*(x)\cdot,\, \cdot \rangle }$ 
	and its dual local norm $\|\cdot\|_x^*= \sqrt{ \langle [\nabla^2A^*(x)]^{-1}\cdot,\, \cdot \rangle }$,
	\fdk{The second one should be 
	$\sqrt{ \langle [\nabla^2A^*({\color{red} \nabla A(x)})]^{-1}\cdot,\, \cdot \rangle }$?
	}
	\item the distance function $\omega(t) = t-\ln (1+t),\;t\geq 0$, and its dual $\;\omega^*(t) = -t-\ln(1-t)$ defined for $t\in [0,1]$.
\end{itemize}
Note that $\omega^*(t)$ is positive, convex and monotonically increasing for $t\in[0,1]$.

@@

%\begin{proposition}
%	Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then
%	\aligns{
%		\norm{\m^*-\m}_{\mF} < 0.21
%		\implies
%		\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2.
%	}
%\end{proposition}
%\begin{proof}
%	
%\end{proof}

\subsection{Self-concordance and Convex Conjugates}
\begin{proposition}
	\label{prop:conjugateConcordance}
	Let $f:\real \rightarrow \real$ be a self-concordant function of Legendre type. Then its convex conjugate $f^*$ is also self-concordant.
\end{proposition}
\begin{proof}
When $f$ is of Legendre type then its convex conjugate $f^*(y) = \max_x \langle x, y \rangle - f(x)$ has inverse gadients
\begin{align}
    \nabla f \circ \nabla f^*  &= \Id \\
     \nabla f^* \circ \nabla f &= \Id
\end{align}
where $\Id$ is the identity function on the relevant domain. 
Deriving this equality proves that Hessians are inverse of each other 
\begin{align}
    \nabla^2 f(x) \nabla^2 f^*(x^*) = \mI
    \label{eq:conjugateHessian}
\end{align}
when evaluated at conjugate pairs of points $(x,x^*)$ -- e.g. $x^*=\nabla f(x)$ and $x = \nabla f^*(x^*)$.
These Hessians are never $0$.
In 1 dimension, this inequality is more simply written 
\aligns{f''(x) f^{* \prime \prime}(x^*) = 1}
Until now, this should be familiar to the reader.
The interesting surprise is that self-concordance appears when we derive again this equality. 
\begin{align}
    &1= f''(x) f^{* \prime \prime}(f'(x)), \forall x \\
    \implies
    &0 = f'''(x)f^{* \prime \prime}(f'(x))  + f''(x)^2 f^{* \prime \prime \prime}(f'(x))\\
    \implies
    &0 = f'''(x)f^{* \prime \prime}(x^*)  + f''(x)^2 f^{* \prime \prime \prime}(x^*)\\
    \implies
    &0 = \frac{f'''(x)}{f''(x)^{\frac{3}{2}}}  f^{* \prime \prime}(x^*)^{-\half}
    + f''(x)^{\half} \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*)^{\frac{3}{2}}} \\
    \implies
    &0= \frac{f'''(x)}{f''(x)^{\frac{3}{2}}}  + \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*)^{\frac{3}{2}}}
\end{align}
where to reach the last line, we used~\eqref{eq:conjugateHessian} again.
We see that for a pair of conjugate functions, the self-concordance ratios are opposite of each other, which concludes the proof.
\end{proof}

>>>>

\subsection{Multivariate MLE}

In higher dimensions, $X \sim \cN(0, \mu_*)$, $X\in\real^d$, $T(X) = XX^\top$, and the mean parameter $\mu_*$ 
is a $d \times d$ symmetric, positive definite covariance matrix with $p = \frac{d(d+1)}{2}$ degrees of freedom.
Note that here $d$ denotes the dimensionality of the data $X$, rather than the dimensionality of the parameters $\mu_*$.

@@

 	&< p \paren{ \inv{n-d-1} - \inv{2n}}  + \frac{d + 1}{4n(n - d - 1)}\\

@@

 	&= p \frac{n + d +1}{2 n (n-d-1)}  + \frac{d(d + 1)/2}{2dn(n - d - 1)}\\

@@

where $p = \frac{d(d+1)}{2}$ and the last inequality used $\inv{2d}\leq 1$. 

@@

\begin{proof}
	The lower bound can be readily obtained by applying Jensen's inequality 
	to the convex function $x\mapsto\inv{x}$ for $x>0$.
	The upper bound requires more work.
	To start, let us plug in the definition of $\MAPm$
\alignn{
		\expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\nat^*}} 
		= \expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i^2}}  
		= (n_0 + n) \expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} 
		= (n_0 +n) \expect{\inv{a + \chi^2(n)}} \; ,
}
	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and $a=n_0 \frac{\mu_0}{\mu^*}$. The expectation reads
\alignn{
 	\expect{\inv{a + \chi^2(n)}}  
	= \int_0^\infty \inv{a+x} \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{2^{\half[n]}\Gamma(\half[n])} dx \; .
}
	This $1/(a+x)$ is hard to deal with.
	Fortunately, there is a trick to express it as an integral itself , with an exponential that fits well with the $\chi^2$-density,
	\fdk{I don't follow from there.} \rlp{do you have more detailed feedback ? I don't know how to improve it.}
\alignn{
		&\int_1^\infty e^{-\frac{zt}{2}} dt = \frac{2}{z} e^{-\half[z]} \\
		\implies &\inv{z} = \half e^{\half[z]} \int_1^\infty e^{-\frac{zt}{2}} dt \\
		\implies & \boxed{\inv{a+x} = \half e^{\frac{a+x}{2}} \int_1^\infty e^{- \frac{(a+x)t}{2}}dt  \; .}
} 
We can then plug this integral into the expectation and apply Fubini's theorem to get a tractable integral on the inside
\alignn{
		2^{\half[n]}\Gamma(\half[n]) \expect{\inv{a + \chi^2(n)}}  
		&= \int_0^\infty \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{a+x} dx \\
		&= \int_0^\infty dx\  x^{\frac{n}{2}-1}e^{-\half[x]} \half e^{\half[a+x]} \int_1^\infty  dt\  e^{-\half[(a+x)t]}  \\
		&= \half e^{\half[a]} \int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty dx \  x^{\frac{n}{2}-1} e^{-\half[x t]}  \; ,
}
	assuming we can exchange the order of integration.
	%\fdk{Fubini's theorem would be a justification here, right? Are you trying to say ``assuming we can exchange the order of integration''?}\rlp{yes exactly.}
	Now with the change of variable $x = 2\frac{y}{t}$, eg $dx = 2\frac{dy}{t}$, the inner integral becomes
	\begin{align}
		\expect{\inv{a + \chi^2(n)}}  
		&= \frac{e^{\half[a]} }{2^{\half[n]+1}\Gamma(\half[n]) }\int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty 2\frac{dy}{t} \  (2\frac{y}{t})^{\frac{n}{2}-1} e^{-y} \\
		&= \frac{e^{\half[a]} }{2\Gamma(\half[n]) } 
		\underbrace{\int_1^\infty  \frac{e^{-\half[a t]}}{t^{\frac{n}{2}}}  dt}_{E_{\half[n]}(\half[a])  } \ 
		\underbrace{\int_0^\infty  y^{\frac{n}{2}-1} e^{-y} dy}_{\Gamma(\half[n])}
	\end{align}
	so we finally get  the formula valid for all $n\geq 1$
	\begin{align}
		\boxed{\expect{\inv{a + \chi^2(n)}}  = \half e^{\frac{a}{2}} E_{\half[n]}(\half[a])}
		\label{eq:hard_expectation}
	\end{align}
	where the {generalized exponential integral function} is defined as 
	\begin{align}
		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; .
	\end{align}
	Now our goal is to bound this function with simpler functions.
	Fortunately, mathematicians have been working on these integrals for decades.
	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19.E21}{the general bound} \citep[Eq.~8.19.21]{DLMF}
	\begin{align}
		e^x E_k(x) 
		&\leq \inv{x + k - 1} \\
		\iff \expect{\inv{a + \chi^2(n)}}
		&\leq \inv{a + n - 2} \\
		\iff \expect{\frac{\mu^*}{\MAPm}} 
		&\leq \frac{n +n_0}{a + n - 2}
	\end{align}
We are left with a special case when  $n=1$, eg $k=\half$. 
Then we can use the trivial bound 
\begin{align}
	\expect{\inv{a + X^2}} < \inv{a}
	\label{eq:trivial_bound}
\end{align}
to conclude the proof.
\end{proof}

@@

\begin{definition}[Self-concordance]
\label{def:self-concordance}
A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies for all feasible $t$
\alignn{
	\abs{g'''(t)} \leq 2 g''(t)^{\frac{3}{2}} \; .
}
\end{definition}

%\begin{definition}[Pseudo Self-concordance] \citep{bach2010self} 
%A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies 
%\alignn{\abs{g'''(x)} \leq C \norm{v}_2 g''(x)}
%for all feasible $t$ and for some $C\geq 0$.
%\end{definition}

@@

We quickly reviews some important properties of self-concordant functions, introduced in \citep{nesterov2003introductory}. We start with a few notations. Let $A^*$ be a self-concordant function. Then, we write
\begin{itemize}
	\item the local norm $\|\cdot\|_x = \sqrt{ \langle \nabla^2A^*(x)\cdot,\, \cdot \rangle }$ 
	%and its dual local norm $\|\cdot\|_x^*= \sqrt{ \langle [\nabla^2A^*(x)]^{-1}\cdot,\, \cdot \rangle }$,
%	\fdk{The second one should be 
%	$\sqrt{ \langle [\nabla^2A^*({\color{red} \nabla A(x)})]^{-1}\cdot,\, \cdot \rangle }$?
%	}
	\item the distance function $\omega(t) = t-\ln (1+t),\;t\geq 0$, and its dual $\;\omega^*(t) = -t-\ln(1-t)$ defined for $t\in [0,1]$.
\end{itemize}
Note that $\omega^*(t)$ is positive, convex and monotonically increasing for $t\in[0,1]$.

@@
